---
title: "Tree Canopy - COVID death rate"
author: "Clarissa Boyajian"
date: "11/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

options(scipen = 999)

library(tidyverse)
library(lubridate)
library(janitor)
library(sf)
library(tmap)
library(patchwork)
library(broom)
```

```{r Read in data}
county_geom_raw <- st_read("map_data/counties.shp")

income_raw <- read.csv("income_data/UnemploymentReport.csv")

tree_canopy_raw <- read_csv("tree_data/HPI-Individual-Indicators/HPI2_treecanopy.csv")

covid_raw_latimes <- read.csv("covid_data/cdph-county-cases-deaths.csv") # 2020-02-01 to 2021-11-22
```

```{r Parameters}
county_region1 <- c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", 
                    "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", 
                    "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba")
county_region2 <- c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", 
                    "Sonoma", "Trinity")
county_region3 <- c("Alameda", "Contra Costa", "Marin", "San Francisco", 
                    "San Mateo", "Santa Clara", "Solano") 
county_region4 <- c("Amador", "Calaveras", "Madera", "Mariposa", "Merced", 
                     "Mono", "San Joaquin", "Stanislaus", "Tuolumne")
county_region5 <- c("Monterey", "San Benito", "San Luis Obispo", 
                    "Santa Barbara", "Santa Cruz", "Ventura")
county_region6 <- c("Fresno", "Inyo", "Kern", "Kings", "Tulare")
county_region7 <- c("Riverside", "San Bernardino")
county_region8 <- c("Los Angeles")
county_region9 <- c("Orange")
county_region10 <- c("Imperial", "San Diego")
```

## Data Wrangling

```{r Wrangling geometry data}
county_geom <- 
  county_geom_raw %>% 
  clean_names() %>% 
  rename(county_name = name) %>% 
  mutate(county_region = case_when(county_name %in% county_region1 ~ "01",
                                   county_name %in% county_region2 ~ "02",
                                   county_name %in% county_region3 ~ "03",
                                   county_name %in% county_region4 ~ "04",
                                   county_name %in% county_region5 ~ "05",
                                   county_name %in% county_region6 ~ "06",
                                   county_name %in% county_region7 ~ "07",
                                   county_name %in% county_region8 ~ "08",
                                   county_name %in% county_region9 ~ "09",
                                   county_name %in% county_region10 ~ "10"),
         county_region_name = case_when(county_name %in% county_region1 ~ "Superior California",
                                        county_name %in% county_region2 ~ "North Coast",
                                        county_name %in% county_region3 ~ "San Francisco Bay Area",
                                        county_name %in% county_region4 ~ "Northern San Joaquin Valley",
                                        county_name %in% county_region5 ~ "Central Coast",
                                        county_name %in% county_region6 ~ "Southern San Joaquin Valley",
                                        county_name %in% county_region7 ~ "Inland Empire",
                                        county_name %in% county_region8 ~ "Los Angeles County",
                                        county_name %in% county_region9 ~ "Orange County",
                                        county_name %in% county_region10 ~ "San Diego - Imperial")) %>% 
  mutate(county_region_name = fct_relevel(county_region_name, 
                                          levels = c("Superior California", "North Coast", 
                                                     "San Francisco Bay Area", "Northern San Joaquin Valley", 
                                                     "Central Coast", "Southern San Joaquin Valley", "Inland Empire", 
                                                     "Los Angeles County", "Orange County", "San Diego - Imperial")),
         county_fips = paste0("06", county_fips)) %>% 
  select(county_fips, county_name, geometry, county_region, county_region_name)
```

```{r Wrangle income data}
income_data <- 
  income_raw %>% 
  clean_names() %>% 
  rename(county_fips = fips) %>% 
  filter(!grepl(c("06000"), county_fips), # remove CA
         !grepl(c("06003"), county_fips), # remove county not in tree data
         grepl(c(" "), name) # remove empty lines at end of csv
         ) %>% 
  select(county_fips, median_household_income_2019)
```

```{r Wrangling tree data}
tree_data <- 
  tree_canopy_raw %>% 
  clean_names() %>% 
  select(-c(indicator, treecanopy_n, treecanopy_d, notes) # remove empty columns
         ) %>% 
  mutate(county_fips = paste0("0", county_fips))
```

```{r Wrangling covid data}
covid_data <- 
  covid_raw_latimes %>% 
  clean_names() %>% 
  filter(!grepl(c("Alpine"), county)) %>% # remove county not in tree data
  mutate(fips = case_when(fips < 10 ~ paste0("0600", fips),
                          fips < 100 & fips > 10 ~ paste0("060", fips),
                          fips > 100 ~ paste0("06", fips))) %>%
  rename(county_name = county,
         county_fips = fips) %>% 
  mutate(date = as.Date(ymd(date)))
```


## Analysis

```{r County average covid data}
covid_county_average <- 
  covid_data %>% 
  group_by(county_fips) %>% 
  summarise(population_av = mean(population),
            confirmed_cases_av = mean(confirmed_cases),
            reported_deaths_av = mean(reported_deaths)) %>% 
  mutate(deaths_per_case_pct = reported_deaths_av / confirmed_cases_av,
         deaths_per_pop_pct = reported_deaths_av / population_av)
```

```{r County average tree data}
tree_county_average <- 
  tree_data %>% 
  group_by(county_fips) %>% 
  summarise(county_treecanopy_pct = mean(treecanopy_pct))
```


```{r Combining data}
full_data_county_average <- 
  left_join(x = covid_county_average, y = tree_county_average, by = "county_fips") %>% 
  left_join(y = county_geom, by = "county_fips") %>% 
  left_join(y = income_data, by = "county_fips")

tree_county_average_geom <- 
  left_join(x = tree_county_average, y = county_geom, by = "county_fips") %>% 
  st_as_sf()

covid_county_average_geom <- 
  left_join(x = covid_county_average, y = county_geom, by = "county_fips") %>% 
  st_as_sf()
```


## Map viz

```{r tree map viz, warning=FALSE}
tree_map <- 
  tm_shape(tree_county_average_geom) +
  tm_borders(col = "black", alpha = .65) +
  tm_fill(col = "county_treecanopy_pct",
          style = "cont", 
          n = 5,
          palette ="viridis",
          title = "Tree Canopy Cover\nPercentage") +
  tm_layout(main.title = "Average Tree Canopy Cover in CA Counties",
            main.title.size = 1)

tmap_save(tree_map, "figs/tree_map.png")

tree_map
```

```{r covid map viz, warning=FALSE}
covid_map <- 
  tm_shape(covid_county_average_geom) +
  tm_borders(col = "black", alpha = .65) +
  tm_fill(col = "deaths_per_case_pct",
          style = "cont", 
          n = 5,
          palette ="viridis",
          title = "Percent of Deaths\nper Positive Case") +
  tm_layout(main.title = "Average COVID Death per Positive Case in CA Counties",
            main.title.size = 1)

tmap_save(covid_map, "figs/covid_map.png")

covid_map
```



## Initial Analysis and basic viz

```{r Covid basic analysis}
covid_county_region_month_average <- 
  covid_county_average_geom %>% 
  group_by(county_region_name, county_region) %>% 
  summarise(county_region_monthly_death_av = mean(reported_deaths_av),
            county_region_monthly_positive_test_av = mean(confirmed_cases_av))
```

```{r Tree basic analysis}
tree_county_region <- 
  tree_county_average_geom %>% 
  group_by(county_region_name, county_region) %>% 
  summarise(county_region_tree_pct = mean(county_treecanopy_pct))
```

```{r Plot basic analysis}
covid_county_region_month_av_plot <- 
  ggplot(data = covid_county_region_month_average, 
         aes(x = county_region_monthly_death_av, 
             y = county_region_name, 
             fill = county_region_name)) +
  geom_col(show.legend = FALSE)+
  scale_fill_viridis_d() +
  labs(x = "Average Monthly COVID Deaths\nfrom 2020-02-01 to 2021-11-22",
       y = "") +
  theme_classic()

tree_county_region_plot <- 
  ggplot(data = tree_county_region,
         aes(x = county_region_tree_pct,
             y = county_region_name,
             fill = county_region_name)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(x = "Average Tree Canopy Cover (%)",
       y = "") +
  theme_classic()

comb_county_region_plot <- 
  (tree_county_region_plot / covid_county_region_month_av_plot) + 
  plot_annotation(title = "California County Regions")

ggsave(plot = comb_county_region_plot, filename = "figs/county_region_tree_covid_plot.png")

comb_county_region_plot
```


## Analysis

Null hypothesis: In California counties, the tree canopy cover percentage has no impact on the rate of COVID deaths per positive reported case.

Alternative hypothesis: In California counties, the tree canopy cover percentage has an impact on the rate of COVID deaths per positive reported case.

```{r Calculate regression model}
mod <- lm(deaths_per_case_pct ~ county_treecanopy_pct, data = full_data_county_average)

mod_summary <- summary(mod)
mod_summary
```

```{r Plot regression}
mod_plot <- 
  ggplot(data = full_data_county_average, 
         aes(x = county_treecanopy_pct, 
             y = deaths_per_case_pct)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  labs(x = "Tree Canopy Cover (%)",
       y = "COVID Deaths per Positive Case (%)",
       title = "Impacts of Tree Canopy Cover on COVID Deaths per Postive Case \nin California Counties")

ggsave(plot = mod_plot, filename = "figs/simple_regression_plot.png")

mod_plot
```

```{r 95% Confidence interval}
crit_val <- qt(0.025, df = 55, lower.tail = FALSE)

point_est <- mod_summary$coefficients[2, "Estimate"]

SE <- mod_summary$coefficients[2, "Std. Error"]

ci_lower <- round(point_est - (crit_val * SE), 6)
ci_upper <- round(point_est + (crit_val * SE), 6)
```








**control for income (don't do interaction)
```{r Calculate regression model controlling for income, eval=FALSE}
mod_income <- lm(deaths_per_case_pct ~ county_treecanopy_pct + median_household_income_2019, data = full_data_county_average)

mod_income_summary <- summary(mod_income)
mod_income_summary
```

```{r Plot regression controlled for income, eval=FALSE}
ggplot(data = full_data_county_average, 
       aes(x = county_treecanopy_pct, 
           y = deaths_per_case_pct)) +
  geom_point() +
  geom_smooth(method = "lm")
```




### References
1. 
2. 
